Generating Random Logic Programs Using Constraint Programming

### 2020-01-20 paper/paper.tex
And the data is in (with a nice plot)

\begin{definition}
  The body of a clause is defined by:
  \begin{itemize}
  \item $\arrayd{cstructure}{\maxNumNodes{}}{[0, \maxNumNodes{} - 1]}$ such
    that:
    \begin{itemize}
    \item $\variable{structure}[i] = i$: the $i$-th node is a root.
    \item $\variable{structure}[i] = j$: the $i$-th node's parent is node $j$.
    \end{itemize}
  \item $\arrayt{values}{\maxNumNodes{}}{Node}$.
  \end{itemize}
\end{definition}

@@

\begin{constraint} \label{constraint:node_types}
  For $i = 0, \dots, \maxNumNodes{} - 1$, let $C_i$ be the number of times $i$
  appears in the \variable{struture} array with index greater than $i$. Then
  \begin{align*}
    C = 0 &\iff \variable{values}[i].\variable{name} \in \predicates{} \cup \{ \top \},\\
    C = 1 &\iff \variable{values}[i].\variable{name} = \neg,\\
    C > 1 &\iff \variable{values}[i].\variable{name} \in \{ \land, \lor \},
  \end{align*}
  where $\variable{structure}_{>i}$ denotes array $\variable{structure}$ with positions
  greater than $i$.
\end{constraint}
Each of the three equivalences in \cref{constraint:node_types} corresponds to
node $i$ having no children, one child, and multiple children, respectively.

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \node[draw,circle,gray,text=black] (and) at (0, 0) {$\land$};
    \node[draw,circle,gray,text=black] (not) at (1.5, -1) {$\neg$};
    \node[draw,gray,text=black] (parent) at (-1.5, -1) {$\mathsf{parent}(X, Y)$};
    \node[draw,gray,text=black] (mother) at (1.5, -2) {$\mathsf{mother}(X, Y)$};
    \draw[gray] (and) -- (parent);
    \draw[gray] (and) -- (not);
    \draw[gray] (not) -- (mother);
  \end{tikzpicture}
  \caption{The tree representation of the body of Clause~\eqref{eq:example_clause}.}
  \label{fig:example_tree2}
\end{figure}
\begin{figure}
  \centering
  \begin{tikzpicture}
    \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
    \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
    \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
    \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
    \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (sibling);
    \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (father);
    \draw[-{Stealth[scale=1.5]}] (mother) edge node[below] {$-$} (father);
  \end{tikzpicture}
  \caption{The predicate dependency graph of the program in \cref{ex:program}.
    Positive edges are labeled with `$+$', and negative edges with `$-$'.}
  \label{fig:predicate_dependencies}
\end{figure}

@@

@@

@@

% TODO: make sure my definitions and examples are consistent with loops

% TODO: Discuss that the algorithms work in case there are multiple 'paths'
% without any issues.

@@

\begin{algorithm}[t]
  \KwData{a program $\mathscr{P}$}
  \SetKwFunction{hasNegativeCycles}{hasNegativeCycles}
  Let $\mathscr{R} \subseteq \mathscr{P}$ be the largest subprogram of
  $\mathscr{P}$ with its $\variable{structure}$ and predicates in both body
  and head fully determined\footnotemark\;
  \lIf{$\mathscr{R} = \emptyset$}{\Return{\textsc{undefined}}}
  \If{\hasNegativeCycles($G_{\mathscr{R}}$)}{
    \Return{\textsc{false}}\;
  }
  \lIf{$\mathscr{R} = \mathscr{P}$}{\Return{\textsc{true}}}
  \Return{\textsc{undefined}}\;
  \caption{Entailment check for negative cycles}
  \label{alg:negative_cycles}
\end{algorithm}
\footnotetext{The arguments (whether variables or constants) are irrelevant to
  our definition of independence.}

@@

\section{EMPIRICAL PERFORMANCE}
% TODO: describe the experimental setup (variables), hardware/software (CPU,
% anything else?)

@@

We also employ a geometric restart policy, restarting after 10, 20, 40,...
failures (contradictions).

% TODO: consider replacing \maxNumClauses{} with max_num_additional_clauses
\begin{itemize}
\item $|\predicates{}|$, $|\variables{}|$, $|\constants{}|$, $\maxNumNodes{}$,
  the number of additional clauses are in $\{ 1, 2, 4, 8, 16 \}$.
\item $\maxArity{} \in \{ 1, 2, 3, 4, 5 \}$.
\item all possible numbers of independent pairs: $0, \dots,
  \binom{|\predicates{}|}{2}$.
\end{itemize}
Repeat 10 times:
\begin{itemize}
\item Pick a random combination of $|\predicates{}|$ arities such that
  $\maxArity{}$ is occurs at least once (in an ordered way, since order doesn't
  matter).
\item Pick the required number of independent pairs (without repetition).
\item Run the solver with a \SI{60}{\second} timeout.
\end{itemize}

Results:
\begin{itemize}
\item ??\% of total running times is less than 1 s
\item ??\% timed out. All of them had $\maxNumNodes{} = 16$.
\item Out of the ten almost-identical runs with each set of parameters, the
  number of runs that timed out was always less than ??.
\item Running time (or number of nodes) as a function of various parameters.
\end{itemize}

@@

\begin{itemize}
\item A constraint for logical equivalence. An algorithm to reduce each tree to
  some kind of normal form. Not doing this on purpose. Leaving for further work.
\item Perhaps negative cycle detection could use the same graph as the
  independence propagator? If we extend each domain to {-1, 0, 1}, but that
  might make propagation weaker or slower.
\item Could investigate how uniform the generated distribution of programs is.
  Distributions of individual parameters will often favour larger values
  because, e.g., there are more 5-tuples than 4-tuples.
\item Inference options to explore. Logspace vs normal space. Symbolic vs
  non-symbolic. Propagate evidence (might be irrelevant)? Propagate weights?
  Supported knowledge compilation techniques: sdd, sddx, bdd, nnf, ddnnf, kbest,
  fsdd, fbdd.
\item Mention the random heuristic. Mention that restarting gives better
  randomness, but duplicates become possible. Restarting after each run is
  expensive. Periodic restarts could be an option.
\item Could add statistics about what constraints tend to conflict.
\end{itemize}

>>>>

\begin{definition}
  The body of a clause is defined by:
  \begin{itemize}
  \item $\arrayd{structure}{\maxNumNodes{}}{[0, \maxNumNodes{} - 1]}$ such
    that:
    \begin{itemize}
    \item $\variable{structure}[i] = i$: the $i$-th node is a root.
    \item $\variable{structure}[i] = j$: the $i$-th node's parent is node $j$.
    \end{itemize}
  \item $\arrayt{values}{\maxNumNodes{}}{Node}$.
  \end{itemize}
\end{definition}

@@

\begin{constraint} \label{constraint:node_types}
  For $i = 0, \dots, \maxNumNodes{} - 1$, let $C_i$ be the number of times $i$
  appears in the \variable{struture} array with index greater than $i$. Then
  \begin{align*}
    C_i = 0 &\iff \variable{values}[i].\variable{name} \in \predicates{} \cup \{ \top \},\\
    C_i = 1 &\iff \variable{values}[i].\variable{name} = \neg,\\
    C_i > 1 &\iff \variable{values}[i].\variable{name} \in \{ \land, \lor \}.
  \end{align*}
\end{constraint}
Each of the three equivalences in \cref{constraint:node_types} corresponds to
node $i$ having no children, one child, and multiple children, respectively.

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node[draw,circle,gray,text=black] (and) at (0, 0) {$\land$};
    \node[draw,circle,gray,text=black] (not) at (1.5, -1) {$\neg$};
    \node[draw,gray,text=black] (parent) at (-1.5, -1) {$\mathsf{parent}(X, Y)$};
    \node[draw,gray,text=black] (mother) at (1.5, -2) {$\mathsf{mother}(X, Y)$};
    \draw[gray] (and) -- (parent);
    \draw[gray] (and) -- (not);
    \draw[gray] (not) -- (mother);
  \end{tikzpicture}
  \caption{The tree representation of the body of Clause~\eqref{eq:example_clause}.}
  \label{fig:example_tree2}
\end{figure}
\begin{figure}
  \centering
  \begin{tikzpicture}
    \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
    \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
    \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
    \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
    \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (sibling);
    \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (father);
    \draw[-{Stealth[scale=1.5]}] (mother) edge node[below] {$-$} (father);
  \end{tikzpicture}
  \caption{The predicate dependency graph of the program in \cref{ex:program}.
    Positive edges are labeled with `$+$', and negative edges with `$-$'.}
  \label{fig:predicate_dependencies}
\end{figure}

@@

  \label{alg:independence_propagation}

@@

  \label{alg:dependencies}

@@

% TODO: start here
% TODO: make sure my definitions and examples are consistent with loops
% TODO: Discuss that the algorithms work in case there are multiple 'paths'
% without any issues.

Note that if there are multiple paths in the dependency graph from $\mathsf{Q}$
to $\mathsf{P}$, when constructing $D_{\mathsf{P}}$, \cref{alg:dependencies}
could include $\mathsf{Q}$ once for each possible type ($\Determined$,
$\Undetermined$, and $\AlmostDetermined$). However,
\cref{alg:independence_propagation,alg:independence_entailment}

@@

\begin{algorithm}
  \KwData{a program $\mathscr{P}$}
  \SetKwFunction{hasNegativeCycles}{hasNegativeCycles}
  Let $\mathscr{R} \subseteq \mathscr{P}$ be the largest subprogram of
  $\mathscr{P}$ with its $\variable{structure}$ and predicates in both body
  and head fully determined\footnotemark\;
  \lIf{$\mathscr{R} = \emptyset$}{\Return{\textsc{undefined}}}
  \If{\hasNegativeCycles($G_{\mathscr{R}}$)}{
    \Return{\textsc{false}}\;
  }
  \lIf{$\mathscr{R} = \mathscr{P}$}{\Return{\textsc{true}}}
  \Return{\textsc{undefined}}\;
  \caption{Entailment check for negative cycles}
  \label{alg:negative_cycles}
\end{algorithm}
\footnotetext{The arguments (whether variables or constants) are irrelevant to
  our definition of independence.}

@@

\section{EMPIRICAL PERFORMANCE}
% TODO: describe the experimental setup (variables), hardware/software (CPU, anything else?)
% see Downloads/cp2019-wiggles.pdf for an example.

\begin{figure*}[t]
  \centering
  \input{phase_transition.tex}
  \caption{TODO}
\end{figure*}

@@

We also employ a geometric restart policy, restarting after $10, 20, 40,\dots$
failures (contradictions).

% TODO: consider replacing \maxNumClauses{} with max_num_additional_clauses
\begin{itemize}
\item $|\predicates{}|$, $|\variables{}|$, $|\constants{}|$, $\maxNumNodes{}$,
  the number of additional clauses are in $\{ 1, 2, 4, 8 \}$.
\item $\maxArity{} \in \{ 1, 2, 3, 4 \}$.
\item all possible numbers of independent pairs: $0, \dots,
  \binom{|\predicates{}|}{2}$.
\end{itemize}
Repeat 10 times:
\begin{itemize}
\item Pick a random combination of $|\predicates{}|$ arities such that
  $\maxArity{}$ is occurs at least once (in an ordered way, since order doesn't
  matter).
\item Pick the required number of independent pairs (without repetition).
\item Run the solver with a \SI{60}{\second} timeout.
\end{itemize}

Results:
\begin{itemize}
\item almost \num{400000} experiments.
\item \SI{97.7}{\percent} of total running times is less than \SI{1}{\second}
\item 4 timeouts: all with 8 predicates, 8 additional clauses, and
  $\maxNumNodes{} = 8$; the rest of the parameters differ.
\item Running time (or number of nodes) as a function of various parameters:
  \begin{itemize}
  \item The number of clauses has a clear impact on time but not number of
    nodes.
  \item $\maxNumNodes{}$ has a huge impact on both.
  \item The proportion of independent pairs of predicates is acting weird.
    Investigate.
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item A constraint for logical equivalence. An algorithm to reduce each tree to
  some kind of normal form. Not doing this on purpose. Leaving for further work.
\item Could investigate how uniform the generated distribution of programs is.
  Distributions of individual parameters will often favour larger values
  because, e.g., there are more 5-tuples than 4-tuples.
%\item Inference options to explore. Logspace vs normal space. Symbolic vs
%  non-symbolic. Propagate evidence (might be irrelevant)? Propagate weights?
%  Supported knowledge compilation techniques: sdd, sddx, bdd, nnf, ddnnf, kbest,
%  fsdd, fbdd.
\item Mention the random heuristic. Mention that restarting gives better
  randomness, but duplicates become possible.
\item Could add statistics about what constraints tend to conflict.
\end{itemize}
